Finiteness in Cubical Type Theory

### 2020-06-28 sections/finiteness-predicates.tex
Update finiteness-predicates.tex

Any attempt to prove that this type is split enumerable will run into trouble:
our task here, then, is to show that such a proof is impossible.
To do so we will first remember Hedberg's theorem
\cite{hedbergCoherenceTheoremMartinLof1998}:
\begin{theorem}[Hedberg's Theorem]
  Every discrete type is a set.
\end{theorem}
With this theorem in mind, then, if we prove that every split enumerable type is
discrete we will have proven that the circle is not split enumerable, because if
it were it would be discrete and therefore a set, which we know already it is
not.

At this point we will define what we mean by \emph{discrete}.
\begin{definition}[Discrete Types]
  A discrete type is one with decidable equality.
  \begin{agdalisting}
    \ExecuteMetaData[agda/Relation/Nullary/Discrete/Base.tex]{discrete-def}
  \end{agdalisting}
\end{definition}
We also use here a data type called \AgdaDatatype{Dec}.
\begin{definition}[Decidable Types]
  If some proposition is decidable it means that it is proven true or false.
  We can encapsulate this with a data type\footnotemark:
  \begin{agdalisting}
    \ExecuteMetaData[agda/Snippets/Dec.tex]{dec-def}
  \end{agdalisting}
  A proof of type \(\AgdaDatatype{Dec}\;A\) is a proof that \(A\) is either true
  or false.
\end{definition}
\footnotetext{This is not, as it happens, the definition of
  \(\AgdaDatatype{Dec}\) you will find in Agda's standard library.
The version of the type that exists there is a good deal more complicated (for
computational reasons),
although its interface behaves}

\begin{definition}[Injections]
  Injective functions are more straightforward to define constructively than
  surjective ones:
  \begin{agdalisting}
    \ExecuteMetaData[agda/Function/Injective/Base.tex]{injective}
  \end{agdalisting} \vspace{-1.5\baselineskip}
  \begin{agdalisting}
    \ExecuteMetaData[agda/Function/Injective/Base.tex]{injection}
  \end{agdalisting}
\end{definition}
\begin{lemma} \label{split-surj-to-inj}
  A split-surjection from \(A\) to \(B\) implies an injection from \(B\) to
  \(A\).
  \begin{agdalisting}
    \ExecuteMetaData[agda/Function/Surjective/Properties.tex]{surj-to-inj}
  \end{agdalisting}
\end{lemma}
\begin{proof} \let\qed\relax

@@

\begin{lemma} \label{split-enum-discrete}
  Every split enumerable type is discrete.
\end{lemma}
\begin{proof}
  Let \(A\) be a split enumerable type.
  By lemma~\ref{split-enum-is-split-surj}, there is a surjection from
  \(\AgdaDatatype{Fin}\;n\) for some \(n\).
  Also, we know that \(\AgdaDatatype{Fin}\;n\) is discrete (proven in our
  formalisation).
  Therefore, by lemma~\ref{discrete-surj}, \(A\) is discrete.
\end{proof}

\section{Manifest Bishop Finiteness} \label{manifest-bishop-finiteness}
We mentioned in the introduction that occasionally in constructive mathematics
proofs will contain ``too much'' information.
With split enumerability we can see an instance of this.
Consider the following proof of the finiteness of the operators for countdown:
\begin{agdalisting} \label{op-slop}
  \ExecuteMetaData[agda/Countdown.tex]{op-slop}
\end{agdalisting}
While it represents the ``same'' information as the proof in
equation~\ref{op-fin}, it clearly is not the same \emph{object}.

>>>>

Any attempt to prove that this type is split enumerable will run into trouble:
our task here, then, is to show that such a proof is impossible.
To do so we will first remember Hedberg's theorem
\cite{hedbergCoherenceTheoremMartinLof1998}:
\begin{theorem}[Hedberg's Theorem]
  Every discrete type is a set.
\end{theorem}
We will in a moment explain what ``discrete'' means, but first we should notice
that his theorem gives us a possible avenue to prove that the circle is not
split enumerable.
We know, for instance, that the circle is \emph{not} a set\todo{proof/reference
  for this?}.
If we can show, then, that every split enumerable type is discrete, we will
have excluded the circle (and indeed any other non-set type).

Of course this doesn't make much sense without defining discrete: it turns out
it's just HoTT-specific language for a concept more usually known as ``decidable
equality''.
\begin{definition}[Discrete Types]
  A discrete type is one with decidable equality.
  \begin{agdalisting}
    \ExecuteMetaData[agda/Relation/Nullary/Discrete/Base.tex]{discrete-def}
  \end{agdalisting}
\end{definition}
Of course we also will need to explain what we mean by ``decidable''.
\begin{definition}[Decidable Types]
  If some proposition is decidable it means that it is proven true or false.
  We can encapsulate this with a data type\footnotemark:
  \begin{agdalisting}
    \ExecuteMetaData[agda/Snippets/Dec.tex]{dec-def}
  \end{agdalisting}
  A proof of type \(\AgdaDatatype{Dec}\;A\) is a proof that \(A\) is either true
  or false.
\end{definition}
\footnotetext{This is not, as it happens, the definition of
  \(\AgdaDatatype{Dec}\) you will find in Agda's standard library.
The version of the type that exists there is a good deal more complicated (for
computational reasons),
although its interface behaves}

So now we have our task in focus: we need to prove that split enumerability
implies decidable equality.
We will do that by first describing split enumerability in terms of injections.
\todo{This proof could be presented a lot better, I think}
\begin{definition}[Injections]
  Injective functions are more straightforward to define constructively than
  surjective ones:
  \begin{agdalisting}
    \ExecuteMetaData[agda/Function/Injective/Base.tex]{injective}
  \end{agdalisting} \vspace{-1.5\baselineskip}
  \begin{agdalisting}
    \ExecuteMetaData[agda/Function/Injective/Base.tex]{injection}
  \end{agdalisting}
\end{definition}
\begin{lemma} \label{split-surj-to-inj}
  A split-surjection from \(A\) to \(B\) implies an injection from \(B\) to
  \(A\).
  \begin{agdalisting}
    \ExecuteMetaData[agda/Function/Surjective/Properties.tex]{surj-to-inj}
  \end{agdalisting}
\end{lemma}
\begin{proof} \let\qed\relax

@@

\begin{lemma} \label{split-enum-discrete}
  Every split enumerable type is discrete.
\end{lemma}
\begin{proof}
  Let \(A\) be a split enumerable type.
  By lemma~\ref{split-enum-is-split-surj}, there is a surjection from
  \(\AgdaDatatype{Fin}\;n\) for some \(n\).
  Also, we know that \(\AgdaDatatype{Fin}\;n\) is discrete (proven in our
  formalisation).
  Therefore, by lemma~\ref{discrete-surj}, \(A\) is discrete.
\end{proof}
\section{Manifest Bishop Finiteness} \label{manifest-bishop-finiteness}
We mentioned in the introduction that occasionally in constructive mathematics
proofs will contain ``too much'' information.
With split enumerability we can see an instance of this.
Consider the following proof of the finiteness of the operators for countdown:
\begin{agdalisting} \label{op-slop}
  \ExecuteMetaData[agda/Countdown.tex]{op-slop}
\end{agdalisting}
While it represents the ``same'' information as the proof in
equation~\ref{op-fin}, it clearly is not the same \emph{object}.
